Type Error Telemetry at Scale: Pseudonymized and Stochiastic

### 2023-09-20 programming23/paper.tex
math mode
draft rest of figure prose

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\stddev}[1]{[#1\textsc{k}]}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1\vphantom{Ap}}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1\vphantom{Ap}}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1\vphantom{Ap}}}

@@

The state of the world in a {Roblox} experience is captured by
the \emph{data model}, which is a tree of \emph{instances}, such as
parts, models, meshes, effects, lighting, audio assets, and physics
constraints such as forces, springs and joints.
The data model can be seen in the Explorer view of Fig.~\ref{fig:roblox-studio}.
While an experience is under development, it is typical for the data
model to be edited (for example instances to be added, deleted, moved
or renamed). Since the initial shape of the data model tree is reflected in
the type system, it is possible for these edits to introduce type errors.

@@

\Cref{f:error-density} presents a zoomed-in view of type error density as it
changes over time in every session of the dataset.
The x-axis counts time in seconds from the start of the session.
The y-axis shows the change in density from the previous telemetry record
in the session to the next.
Changes can be positive or negative depending on whether the edits added
or removed errors.
The plots cut off at \pm{}50 density and 150 seconds.to focus on the
bulk of the data.
%% Sadly even with the zoom, it's too hard to follow one session ... the plot only gives a big-picture impression.

@@

\subsection{Changes to Errors by Mode}

\begin{figure}[t]\centering
  % TODO use bars instead, percent only, don't care about counts
  \begin{tabular}{lr@{}rr@{}rr@{}r}
    & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
    \mnocheck{} & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
    \mnonstrict{} & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
    \mstrict{} & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
  \end{tabular}
  \caption{How often do edits increase, maintain, or decrease the total number of type errors?}
  \label{f:error-changes}
\end{figure}

\FILL{}

Fifth,
Increases and decreases to the number of type errors are evenly balanced across
the data.
In strict mode, \pct{24} of records increase the number of type errors. \pct{26}
decrease. \pct{50} keep it constant.
In nonstrict mode, \pct{14} increase errors, \pct{15} decrease, and \pct{71}
keep it constant
This plot is for nonstrict. X-axis = time since start of session, cut off at 30
seconds. Y-axis = change in num errors since last record, cut off at +/- 100.

\subsection{Data Model}

\begin{figure}[t]\centering
  %% TODO
  %% - thin bars
  %% - colors for good , bad (disagree) , neutral
  \begin{tabular}{lll}
    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
  \end{tabular}
  \caption{FILL Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
  \label{f:tefs}
\end{figure}

\FILL{}

% -\subsection{Asset Creation vs Scripting}
% -
% -Data model. Point to discussion above.
% -
% -Big fluctuations in errors.
% -FS disagrees with TE.
% -Probably data model!

Sixth, DM story.
FS has: 1 all dependencies in strict, 2 data-model awareness (DM turns into
type graph, ... initial DM, all properties go into a table type; DM -> folder
-> model -> part .. get dot-driven autocomplete) in strict, DM might have type
Any might have type Instance, the latter would mean more errors than FS

DM insensitivity in strict mode
(if DM has top type that would lead to errors.)
Context: each of the 3 modes runs 2 kinds of type analysis. One is the regular
type analysis. The other is Forced Strict (FS), which converts all dependencies
to strict mode. The reason for FS is that it sometimes gives better
autocomplete suggestions etc.
You'd expect that the number of type errors is always <= num FS errors. But we
found several cases where this is false!
The plot here "stacks" every strict session into bars. X-axis is a record index
within a session (bars are higher on the left because we have lots more short
sessions than long ones). Y-axis counts the number of records where num TE <= num FS
is true (blue) and false (yellow).

\Cref{f:tefs}~FILL story is edits to DM vs edits to code.
Data model awareness, call for future work.

>>>>

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\stddev}[1]{[#1\textsc{k}]}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1\vphantom{Ap}}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1\vphantom{Ap}}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1\vphantom{Ap}}}
\newcommand{\tefsfont}[1]{\textsc{#1}}
\newcommand{\tekey}{\tefsfont{te}}
\newcommand{\fskey}{\tefsfont{fs}}

@@

The state of the world in a {Roblox} experience is captured by
the \emph{data model}, which is a tree of \emph{instances}, such as
parts, models, meshes, effects, lighting, audio assets, and physics
constraints such as forces, springs and joints.
The data model can be seen in the Explorer view of~\Cref{fig:roblox-studio}.
While an experience is under development, it is typical for the data
model to be edited (for example instances to be added, deleted, moved
or renamed). Since the initial shape of the data model tree is reflected in
the type system, it is possible for these edits to introduce type errors.

@@

\Cref{f:error-density} presents a zoomed-in view of type error density as it
changes over time in every session of the dataset.
The x-axis counts time in seconds from the start of the session.
The y-axis shows the change in density from the previous telemetry record
in the session to the next.
Changes can be positive or negative depending on whether the edits added
or removed errors.
The plots cut off at$\pm{}50$ density and 150 seconds.to focus on the
bulk of the data.
%% Sadly even with the zoom, it's too hard to follow one session ... the plot only gives a big-picture impression.

@@

\subsection{Do Edits Tend to Add Type Errors?}

\begin{figure}[t]\centering
  % TODO use bars instead, percent only, don't care about counts
  %% TODO old version had +-20% add/drop and a bigger % of keeps. What changed?
  \begin{tabular}{lr@{}rr@{}rr@{}r}
    & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
    \mnocheck{} & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
    \mnonstrict{} & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
    \mstrict{} & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
  \end{tabular}
  \caption{How often do edits increase, maintain, or decrease the number of type errors?}
  \label{f:error-changes}
\end{figure}

The apparent visual symmetry in the density plots~(\cref{f:error-density})
suggests that edits add and remove errors with rougly equal frequency,
regardless of mode.
No matter whether a creator is using \mnocheck{}, \mnonstrict{},
or \mstrict{} mode, edits seem to balance increases and decreases in the number
of type errors.
\Cref{f:error-changes} explores this question of balance in detail.
For each of the three modes, it reports the percent of edits that increase (Add),
maintain at a nonzero level (Keep), or decrease (Drop) the number of type
errors in the project.

\begin{itemize}
  \item
    The Adds and Drops are remarkably balanced: across all three modes, they are
    at most \pct{2} apart from one another.
    This confirms that changes to the number of errors are balanced.
    Creators tend to fix errors highlighted by type analysis
    even in the stricter analysis modes.

  \item
    The percentage of Keeps is somewhat high: nearly \pct{10} in \mnocheck{}
    and \pct{20} in \mnonstrict{} and \mstrict{}.
    Evidently, the mode plays a role here.
    Outside of \mnocheck{}, type errors are more likely to persist through edits.

\subsection{Two Activities: Asset Creation and Scripting}

\begin{figure}[t]\centering
  %% TODO
  %% - thin bars
  %% - colors for good , bad (disagree) , neutral
  \begin{tabular}{lll}
    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
  \end{tabular}
  \caption{FILL Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
  \label{f:tefs-compass}
\end{figure}

Since Roblox Studio runs each codebase through two passes of type analysis,
it is interesting to compare their results.
Of special interest is the \mstrict{} mode case, because the number of
type errors and forced strict errors should be nearly the same.
But, there are two differences between \mstrict{} type checking
and forced strict checking:
\begin{itemize}
  \item
    Forced strict checks all dependencies in \mstrict{} mode, no matter
    what mode they declare.
    So, the number of forced strict errors may be higher.
  \item
    Forced strict has looser rules for the \emph{data model};
    that is, for the static assets that a program manipulates.
    So, the number of forced strict errors may be lower.
    %% initial DM, all properties go into a table type; DM -> folder -> model -> part
    %% get dot-driven autocomplete
\end{itemize}
Our data contains a surprising number of records where forced strict
reports fewer errors.
Data model edits are apparently a regular occurrence in Luau programs.

\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in forced
strict errors (\fskey{}) for the three analysis modes.
To a first approximation, the changes either agree with one another
or disagree.
For example, if type errors increase but forced strict errors decrease,
we have a disagreement.
The plots futher categorize agreements and disagreements into
the following five cases:
\begin{enumerate}
  \item
    \emph{zero}: neither \tekey{} nor \fskey{} changed
  \item
    \emph{te only}: \tekey{} changed but \fskey{} remained the same
  \item 
    \emph{fs only}: \tekey{} remained the same but \fskey{} changed
  \item
    \emph{dis}: \tekey{} and \fskey{} changed in opposite directions
  \item 
    \emph{agree}: \tekey{} and \fskey{} changed in the same direction
\end{enumerate}
Both \emph{zero} and \emph{agree} are agreements.
The others are disagreements.
The plots show what percent of all telemetry records fell into these five categories.

\begin{itemize}
  \item
    All three plots have a nonzero percentage of \emph{te only} items.
    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
    Since \fskey{} is more lenient only for data model changes, these must
    be edits that affect the model in a way that raises a type error.
    Naturally, \mstrict{} mode has the largest number of such edits because
    its type checker is the most careful.

  \item
    In the first plot, for \mnocheck{},
    half the records are \emph{zero\/}s, nearly half are \emph{fs only}, and only a few
    land in the other categories.
    While the number of zeros is a bit high,
    the \emph{fs. only} spike makes sense because type checking is supposed to complain
    sparingly in \mnocheck{} mode.

  \item
    In the \mnonstrict{} plot, the number of agreements jumps up.
    This is reassuring.
    \FILL{}.

Users of \mstrict{} mode may find it frustrating that the type checker complains
about data model edits.
Though we cannot see the exact reason for the complaints, the checker errs on
the conservative side.
Thus we expect the number of useful reports is low.
This may be a barrier to adaption.
